Computational biology in the 21st century:\\Algorithms that scale\\Supplement

### 2015-07-31 main.tex
final tweaks for submission

\title{Computational biology in the 21st century: Algorithms that scale}
\numberofauthors{3}

@@

This growth in data poses significant challenges for 
researchers~\cite{marx2013biology}.
Currently, many biological `omics' applications require us to store, access, 
and analyze large libraries of data.
One approach to solving these challenges is to embrace cloud computing.
Google, Inc. and the Broad Institute have collaborated to bring the GATK 
(Genome 
Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
Amazon Web Services are also commonly used for computational biology 
research and enterprise (e.g., DNANexus)~\cite{schatz2010cloud}.
However, while cloud computing frees researchers from maintaining their own
data centers and provides cost-saving benefits when computing resources are
not needed continuously, it is no panacea.
First and foremost, the computer systems that make up those cloud datacenters
are themselves bound by improvements in semiconductor technology and Moore's
law. Thus, cloud computing does not truly address the problem posed by the
faster-than-Moore's law exponential growth in omics data.
Moreover, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
in West Africa, analysis resources are needed at often-remote field sites.
While it is now possible to bring sequencing equipment and limited computing 
resources to remote sites, internet connectivity is still highly constrained.
Thus, accessing cloud resources for analytics may not be possible.

Computer scientists routinely exploit the structure of various data in
order to reduce time or space complexity.
In computational biology, this approach has implicitly served researchers well.
Now-classical approaches such as principal component analysis (PCA) reduce the 
dimensionality of data in order to simplify analysis and uncover salient 
features~\cite{berger2013computational}.
As another example, clever indexing techniques such as the Burroughs-Wheeler 
Transform (BWT) take advantage of aspects of sequence 
structure~\cite{berger2013computational} to speed up computation and save storage.
This review focuses on cutting-edge algorithmic advances for dealing with the growth in 
biological data by explicitly taking advantage of its unique structure; algorithms for gaining novel biological insights are not its 
focus.

@@

% YWY: "Most" is an extremely strong claim. I believe you, but I could see people
% arguing (e.g. all of electronic health records could be considered "biological")
Most but not all biological data take the form of \emph{sequence} data, either 
nucleotide sequences (using a four-letter alphabet representing the four 
DNA or RNA bases) or protein sequences (using a twenty-letter alphabet 
representing the twenty standard amino acids).
Sequence data are obtained in several ways, including mass spectrometry (which
can determine protein sequence, as well as interactions) and RNA-seq (which can determine RNA sequence and quantity, 
allowing scientists to also infer the expression of the gene to which it might translate).
However, the greatest volume of sequence data available are genomic in nature and 
come from next-generation sequencing (NGS) technologies.

@@

While sequence is important, scientists are also interested in measuring gene expression, or the quantity of
a particular gene product (usually a protein or more recently RNA) present in the cell.
Gene expression data is useful for relating genotype to phenotype, and is 
important in studying many diseases.
Usually obtained through microarrays or RNA-seq technologies, expression data is
quantitative, as each gene from a sample is associated with a numeric expression
level, and high-dimensional, as many thousands of genes may be analyzed at a
time.
Resources such as NCBI’s Gene Expression Omnibus (GEO), which pool 
together many diverse expression studies, now allow researchers to analyze 
thousands of samples generated by others to obtain meaningful biological 
insights. 
More importantly, many of these biological insights can only be gleaned when 
viewed in the context of hundreds or thousands of gene expression samples.

Chemical compounds represent a source of structural data.
Determined by X-ray crystallography, NMR, electron microscopy and other 
techniques, chemical compounds---many of which are useful pharmaceuticals---can be
represented as labeled graphs (or equivalently, adjacency matrices).
In such graphs, the vertices are atoms (labeled with a chemical element) and
the edges are bonds between atoms (labeled with a bond type, such as single or
double).
Chemical compounds are stored in databases such as PubChem, and are often searched
for by medicinal chemists as potential agonists for protein targets.

@@

Given DNA or RNA reads from NGS technologies, the first task is to 
\emph{assemble} those fragments of sequence (which have associated quality 
scores) into contiguous sequences.
The task of \emph{de novo} assembly is beyond the scope of this Review, but is
possible because reads overlap and because the sequence is covered by many such
overlapping reads~\cite{zhang2011practical}; the \emph{de Bruijn graph} is a 
commonly-used data structure.
Often, however, a \emph{reference genome} (or in the case of RNA, 
\emph{transcriptome}) is available for the organism being sequenced; the
establishment of a human reference genome was indeed the purpose of the Human
Genome Project.

@@

Driven by the plummeting costs of next-generation sequencing, the 1000 Genomes Project~\cite{10002012integrated} is pursuing a broad catalog of human 
variation; instead of producing a single reference genome for a species, many
complete genomes are catalogued.
Likewise, WormBase~\cite{harris2014wormbase} and 
FlyBase~\cite{tweedie2009flybase} are cataloguing many different species and
strains of the \emph{Caenorhabditis} worm and \emph{Drosophila} fruit fly, 
respectively.
These genomes are enabling cross-species inference, for example about genes and regulatory regions, and thus insights into function and
evolution~\cite{berger2013computational}.
Again, the sheer enormity of sequencing data is problematic for storage, access and analysis.

On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease~\cite{gevers2014treatment}, and 
obesity~\cite{greenblum2012metagenomic}.
Metagenomics fundamentally asks what organisms are present, and, in the case
of a microbiome such as the gut, what metabolic functions it can accomplish as
a whole.
One way of addressing this problem is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second way is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
The difficulty is that the datasets required to shine any light on these 
questions, namely from `shotgun' metagenomics, are gigantic and vastly more 
complex than the 16S metagenomic studies that merely classify the bacteria 
based on 
their divergence with respect to a specific gene (namely, the 16S ribosomal 
subunit).
The massive data results in major identification 
challenges for certain bacterial, as well as viral, species and 
genera~\cite{janda200716s}.

Given a sequenced genome, the next natural questions ask what genes (genomic
regions that code for proteins) are present, what structure each resulting
protein takes, and what biological function it performs.
Identifying likely genes is a well-studied problem~\cite{burge1998finding} 
beyond the scope of this Review.
However, determining evolutionary relationships, structure, and function is
at the heart of current research in computational biology.
Since some organisms (known as \emph{model organisms}) are better studied than
others, and evolution is known to conserve sequence, structure, and function,
a powerful approach to determine these attributes is to search for similar
sequences about which more is known. 
This so-called \emph{homology search} entails searching for \emph{approximate}
matches in databases of known gene or protein sequences.
The homology search problem was previously believed to be solved;
Basic Local Alignment Search Tool (BLAST)~\cite{altschul1990basic} had been
the standard tool for performing homology (similarity) search on databases of nucleotide and
protein sequences.
BLAST takes a `seed-and-extend' approach; it looks for small, $k$-mer matches
that might lead to longer matches, and greedily extends them, ultimately 
producing a \emph{sequence alignment} between a query and each potential 
database hit.
However, BLAST's running-time scales linearly with the size of the database being
searched, which is problematic as sequence databases continue to grow at a
faster rate than Moore's law.

While gene expression data lends itself to cluster analysis and probabilistic 
approaches, the high dimensionality and noise in the data presents significant 
challenges~\cite{nguyen2002tumor}.
Principal Component Analysis has shown promise in reducing the dimensionality
of gene expression 
data~\cite{raychaudhuri2000principal,alter2000singular}.
Such data and its challenges have been the focus of other 
articles~\cite{berger2013computational}, and thus will be only lightly touched 
upon here.

@@

Techniques for reference-based read mapping typically rely on algorithmic 
approaches such as the Burrows-Wheeler transform (BWT)~\cite{burrows1994block}, 
which provides efficient string 
compression through a reversible transformation, while the FM-index data 
structure~\cite{ferragina2000opportunistic} is a compressed substring index, 
based on the BWT, which provides efficient storage as well as fast search.
BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
Bowtie2~\cite{langmead2012fast} mapper further relies on the FM-index for 
efficient mapping of NGS reads.
The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
coupled with dynamic programming in a compressed representation of the 
reference genome, in order to prune the search space 
when mapping reads to a reference genome.
Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2010mrsfast} use an `approximate seed' approach to indexing
the space of possible matches, likewise pruning the search space; however, the
bulk of its runtime is spent on the extend phase.
State-of-the-art mapper mrsFAST-Ultra achieves improvements in efficiency based on machine architecture rather than leveraging redundancy in the data itself with near-perfect sensitivity, but
only for the case where there are no insertions and deletions (indels) \cite{hach2014mrsfast}.
Even with these approaches, read mapping remains a significant bottleneck in
genomic research~\cite{hach2010mrsfast}.

@@

Recent work on gene expession has explored additional ways to exploit the high-dimensional 
structure of the data.
SPARCLE (SPArse ReCovery of Linear combinations of Expression)~\cite{prat2011recovering}
brings ideas from compressed sensing~\cite{candes2005decoding} 
to gene expression analysis.
Specifically, SPARCLE relies on the idea that a randomly-chosen set of points 
are unlikely to be close in a lower-dimensional linear subspace.
Thus, SPARCLE approximates a gene's expression profile as a linear combination
of a few other genes' profiles, and then applies supervised learning to
predict relationships between genes.
Another recent and novel approach to exploiting the structure of gene expression
space is Parti (Pareto task inference)~\cite{hart2015inferring}, which describes a set of
data as a polytope, and infers the specific tasks represented by vertices of
that polytope from the features most highly enriched at those vertices.
In human breast tumors, and in mouse tissues, the expression data were
well-described by tetrahedra whose vertices were enriched for different tumor
types and biological functions, from which four distinct tasks are 
inferred~\cite{hart2015inferring}.

@@

An area of great interest in integrating biological network data is in global multiple network 
alignment, which is based on network topology and conserved biological function (IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn}).
IsoRank's approach is similar to the PageRank algorithm~\cite{page1999pagerank} 
but performs a random walk on the product graph of two networks;
IsoRankN produces a \emph{Personalized PageRank} alignment, similar to the
PageRank-Nibble algorithm~\cite{andersen2006local}.
Diffusion state distance (DSD)~\cite{cao2013going} is a diffusion-based approach
which defines a distance metric between two nodes based on random walks.
DSD penalizes high-degree nodes, and outperforms other graph-theoretic distance
functions when applied to standard approaches such as neighbor voting for function annotation.
Diffusion component analysis (DCA)~\cite{cho2015diffusion} amounts to a network-based 
analogue of principle component analysis; DCA reduces the dimensionality of a
network in such a way that function annotation can be performed in the reduced 
space, significantly decreasing running time while still providing competitive
prediction accuracy.
Using a compressed representation of metabolic networks, 
Mongoose~\cite{chindelevitch2014exact} is able to 
identify and unblock many blocked reactions in current models without loss of 
accuracy.

@@

The CORA read mapper~\cite{yorukoglu2015compressive} applies a mid-size $l$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
CORA, like caBLAST and caBLASTP, accelerates existing tools (in this case, read
mappers including BWA or Bowtie2) by allowing them to operate in a compressed
space, and relying on a coarse and a fine phase.
In contrast, short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
MrsFAST~\cite{hach2010mrsfast}, conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its $l$-mer based read compression model, CORA is able to accelerate 
and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.
Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
not necessarily, a power of two) used as `seeds' for longer sequence matches.
CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each 
one to its neighbors within a small Hamming or Levenshtein distance.
The term $l$-mer distinguishes these substrings from typically-short $k$-mers.

@@

Compressive genomics \cite{loh2012compressive} has been
generalized and adapted to non-sequence spaces as well, and coined
`compressive omics'~\cite{yu2015entropy}.
One such example is chemogenomics.
Applying a compressive acceleration approach, Ammolite~\cite{yu2015entropy}
accelerates SMSD search by an average of 150x on the PubChem database.
Another example is 
esFragBag~\cite{yu2015entropy}, which clusters proteins based on the cosine distance or
Euclidean distance of their bag-of-words vectors, further accelerating FragBag's
running time by an average of 10x.

@@

Along with a genome sequence, modern sequencing systems provide a measure of
the quality of, or confidence in, the nucleotide base read at every position.
While nucleotides reside in a 4-letter alphabet, na\"ively representable in two 
bits per position, these quality scores occupy a larger range, na\"ively
requiring six or seven bits per position.
Thus, quality scores offer low-hanging fruit with respect to compression for 
storage and transmission of sequence data.
Several approaches to compressing quality scores have been 
presented~\cite{janin2013adaptive,bonfield2013compression,ochoa2013qualcomp,yu2015quality,yu2014traversing}.
Quartz~\cite{yu2015quality} takes the unusual approach of applying lossy 
compression, by bounding the likelihood that a quality score is informative.
Importantly, Quartz modifies the FASTQ file of the reads and quality scores, and thus can be
readily incorporated into existing sequence analysis pipelines.
Surprisingly, Quartz not only demonstrates improved compression over competing
approaches, but slightly improves the accuracy of downstream variant-calling.

Also taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
similar reads (those that share a common 
substring) together into `buckets', allowing that common substring to now be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This approach allows a general-purpose compressor to achieve tight compression.
SCALCE~\cite{hach2012scalce} also relies on a `boosting' scheme, reordering
reads in such a way that a general-purpose compressor achieves improved
compression ratios.

@@

\section{Acknowledgements}
This work is supported by the National Institutes of Health, under
grant GM108348.
Y.W.Y. is supported by a Hertz Foundation Fellowship.

>>>>

\title{Computational biology in the 21st century:\\Algorithms that scale}
\numberofauthors{3}

@@

This growth in data poses significant challenges for 
researchers~\cite{marx2013biology}.
Currently, many biological `omics' applications require us to store, access, 
and analyze large libraries of data.
One approach to solving these challenges is to embrace cloud computing.
Google, Inc. and the Broad Institute have collaborated to bring the GATK 
(Genome 
Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
Amazon Web Services are also commonly used for computational biology 
research and enterprise (e.g., DNAnexus)~\cite{schatz2010cloud}.
However, while cloud computing frees researchers from maintaining their own
data centers and provides cost-saving benefits when computing resources are
not needed continuously, it is no panacea.
First and foremost, the computer systems that make up those cloud datacenters
are themselves bound by improvements in semiconductor technology and Moore's
law. Thus, cloud computing does not truly address the problem posed by the
faster-than-Moore's law exponential growth in omics data.
Moreover, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
in West Africa, analysis resources are needed at often-remote field sites.
While it is now possible to bring sequencing equipment and limited computing 
resources to remote sites, internet connectivity is still highly constrained.
Thus, accessing cloud resources for analytics may not be possible.

Computer scientists routinely exploit the structure of various data in
order to reduce time or space complexity.
In computational biology, this approach has implicitly served researchers well.
Now-classical approaches such as principal component analysis (PCA) reduce the 
dimensionality of data in order to simplify analysis and uncover salient 
features~\cite{berger2013computational}.
As another example, clever indexing techniques such as the Burroughs-Wheeler 
Transform (BWT) take advantage of aspects of sequence 
structure~\cite{berger2013computational} to speed up computation and save storage.
This Review focuses on cutting-edge algorithmic advances for dealing with the growth in 
biological data by explicitly taking advantage of its unique structure; algorithms for gaining novel biological insights are not its 
focus.

@@

Much but not all biological data take the form of \emph{sequence} data, either 
nucleotide sequences (using a four-letter alphabet representing the four 
DNA or RNA bases) or protein sequences (using a twenty-letter alphabet 
representing the twenty standard amino acids).
Sequence data are obtained in several ways, including mass spectrometry (which
can determine protein sequence, as well as interactions) and RNA-seq (which can determine RNA sequence and quantity, 
allowing scientists to also infer the expression of the gene to which it might translate).
However, the greatest volume of sequence data available are genomic in nature and 
come from next-generation sequencing (NGS) technologies.

@@

As mentioned earlier, function follows form, so in addition to
sequence and expression, structure plays an important role in biological data science.
Chemical compounds represent a source of structural data.
Determined by X-ray crystallography, NMR, electron microscopy and other 
techniques, chemical compounds---many of which are useful pharmaceuticals---can be
represented as labeled graphs (or equivalently, adjacency matrices).
In such graphs, the vertices are atoms (labeled with a chemical element) and
the edges are bonds between atoms (labeled with a bond type, such as single or
double).
Chemical compounds are stored in databases such as PubChem, and are often searched
for by medicinal chemists as potential agonists for protein targets.

@@

Given DNA or RNA reads from NGS technologies, the first task is to 
\emph{assemble} those fragments of sequence (which have associated quality 
scores) into contiguous sequences.
The task of \emph{de novo} assembly is beyond the scope of this Review, but is
possible because reads overlap and because the sequence is covered by many such
overlapping reads~\cite{berger2013computational}; the \emph{de Bruijn graph} is a 
commonly-used data structure.
Often, however, a \emph{reference genome} (or in the case of RNA, 
\emph{transcriptome}) is available for the organism being sequenced; the
establishment of a human reference genome was indeed the purpose of the Human
Genome Project.

@@

Driven by the plummeting costs of next-generation sequencing, the 1000 Genomes Project~\cite{10002012integrated} is pursuing a broad catalog of human 
variation; instead of producing a single reference genome for a species, many
complete genomes are catalogued.
Likewise, WormBase and 
FlyBase are cataloguing many different species and
strains of the \emph{Caenorhabditis} worm and \emph{Drosophila} fruit fly, 
respectively.
These genomes are enabling cross-species inference, for example about genes and regulatory regions, and thus insights into function and
evolution~\cite{berger2013computational}.
Again, the sheer enormity of sequencing data is problematic for storage, access and analysis.

On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease, and 
obesity~\cite{kostic2014microbiome}.
Metagenomics fundamentally asks what organisms are present, and, in the case
of a microbiome such as the gut, what metabolic functions it can accomplish as
a whole.
One way of addressing this problem is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second way is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
The difficulty is that the datasets required to shine any light on these 
questions, namely from `shotgun' metagenomics, are gigantic and vastly more 
complex than the 16S metagenomic studies that merely classify the bacteria 
based on 
their divergence with respect to a specific gene (namely, the 16S ribosomal 
subunit).
The massive data results in major identification 
challenges for certain bacterial, as well as viral, species and 
genera~\cite{janda200716s}.

Given a sequenced genome, the next natural questions ask what genes (genomic
regions that code for proteins) are present, what structure each resulting
protein takes, and what biological function it performs.
Identifying likely genes is a well-studied problem~\cite{berger2013computational} 
beyond the scope of this Review.
However, determining evolutionary relationships, structure, and function is
at the heart of current research in computational biology.
Since some organisms (known as \emph{model organisms}) are better studied than
others, and evolution is known to conserve sequence, structure, and function,
a powerful approach to determine these attributes is to search for similar
sequences about which more is known. 
This so-called \emph{homology search} entails searching for \emph{approximate}
matches in databases of known gene or protein sequences.
The homology search problem was previously believed to be solved;
Basic Local Alignment Search Tool (BLAST)~\cite{berger2013computational} had been
the standard tool for performing homology (similarity) search on databases of nucleotide and
protein sequences.
BLAST takes a `seed-and-extend' approach; it looks for small, $k$-mer matches
that might lead to longer matches, and greedily extends them, ultimately 
producing a \emph{sequence alignment} between a query and each potential 
database hit.
However, BLAST's running-time scales linearly with the size of the database being
searched, which is problematic as sequence databases continue to grow at a
faster rate than Moore's law.

While gene expression data (see Supplement) lends itself to cluster analysis 
and probabilistic 
approaches, the high dimensionality and noise in the data presents significant 
challenges~\cite{berger2013computational}.
Principal Component Analysis has shown promise in reducing the dimensionality
of gene expression 
data~\cite{berger2013computational}.
Such data and its challenges have been the focus of other 
articles~\cite{berger2013computational}, and thus will be only lightly touched 
upon here.

@@

Techniques for reference-based read mapping typically rely on algorithmic 
approaches such as the Burrows-Wheeler transform (BWT)~\cite{berger2013computational}, 
which provides efficient string 
compression through a reversible transformation, while the FM-index data 
structure~\cite{berger2013computational} is a compressed substring index, 
based on the BWT, which provides efficient storage as well as fast search.
BWA (Burrows-Wheeler Aligner)~\cite{berger2013computational} uses the BWT, while the 
Bowtie2~\cite{berger2013computational} mapper further relies on the FM-index for 
efficient mapping of NGS reads.
The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
coupled with dynamic programming in a compressed representation of the 
reference genome, in order to prune the search space 
when mapping reads to a reference genome.
Masai~\cite{siragusa2013fast} and mrsFAST \cite{berger2013computational} use an `approximate seed' approach to indexing
the space of possible matches, likewise pruning the search space; however, the
bulk of its runtime is spent on the extend phase.
State-of-the-art mapper mrsFAST-Ultra achieves improvements in efficiency based on machine architecture rather than leveraging redundancy in the data itself with near-perfect sensitivity, but
only for the case where there are no insertions and deletions (indels) \cite{hach2014mrsfast}.
Even with these approaches, read mapping remains a significant bottleneck in
genomic research~\cite{berger2013computational}.

@@

Recent work on gene expression has explored additional ways to exploit the high-dimensional 
structure of the data.
SPARCLE (SPArse ReCovery of Linear combinations of Expression)~\cite{prat2011recovering}
brings ideas from compressed sensing~\cite{candes2005decoding} 
to gene expression analysis.
Another recent and novel approach to exploiting the structure of gene expression
space is Parti (Pareto task inference)~\cite{hart2015inferring}, which describes a set of
data as a polytope, and infers the specific tasks represented by vertices of
that polytope from the features most highly enriched at those vertices.

@@

An area of great interest in integrating biological network data is in global multiple network 
alignment, which is based on network topology and conserved biological function (IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn}).
IsoRank's approach is similar to the PageRank algorithm~\cite{berger2013computational} 
but performs a random walk on the product graph of two networks;
IsoRankN produces a \emph{Personalized PageRank} alignment, similar to the
PageRank-Nibble algorithm~\cite{berger2013computational}.
NetworkBLAST~\cite{atias2012comparative} is another approach which searches for 
conserved paths and conserved interaction clusters.
Diffusion state distance (DSD)~\cite{cao2013going} is a diffusion-based approach
which defines a distance metric between two nodes based on random walks.
DSD penalizes high-degree nodes, and outperforms other graph-theoretic distance
functions when applied to standard approaches such as neighbor voting for function annotation.
Diffusion component analysis (DCA)~\cite{cho2015diffusion} amounts to a network-based 
analogue of principle component analysis; DCA reduces the dimensionality of a
network in such a way that function annotation can be performed in the reduced 
space, significantly decreasing running time while still providing competitive
prediction accuracy.
Using a compressed representation of metabolic networks, 
Mongoose~\cite{chindelevitch2014exact} is able to 
identify and unblock many blocked reactions in current models without loss of 
accuracy.

@@

The CORA read mapper~\cite{yorukoglu2015compressive} applies a mid-size $l$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
CORA, like caBLAST and caBLASTP, accelerates existing tools (in this case, read
mappers including BWA or Bowtie2) by allowing them to operate in a compressed
space, and relying on a coarse and a fine phase.
In contrast, short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
MrsFAST~\cite{berger2013computational}, conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its $l$-mer based read compression model, CORA is able to accelerate 
and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.
Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
not necessarily, a power of two) used as `seeds' for longer sequence matches.
CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each 
one to its neighbors within a small Hamming or Levenshtein distance.
The term $l$-mer distinguishes these substrings from typically-short $k$-mers.

@@

Compressive genomics \cite{loh2012compressive} has been
generalized and adapted to non-sequence spaces as well, and coined
`compressive omics.'
One such example is chemogenomics.
Applying a compressive acceleration approach, Ammolite~\cite{yu2015entropy}
accelerates SMSD search by an average of 150x on the PubChem database.
Another example is 
esFragBag~\cite{yu2015entropy}, which clusters proteins based on the cosine distance or
Euclidean distance of their bag-of-words vectors, further accelerating FragBag's
running time by an average of 10x.

@@

Along with a genome sequence, modern sequencing systems provide a measure of
the quality of, or confidence in, the nucleotide base read at every position.
While nucleotides reside in a 4-letter alphabet, na\"ively representable in two 
bits per position, these quality scores occupy a larger range, na\"ively
requiring six or seven bits per position.
Thus, quality scores offer low-hanging fruit with respect to compression for 
storage and transmission of sequence data.
Several approaches to compressing quality scores have been 
presented~\cite{janin2013adaptive,bonfield2013compression,ochoa2013qualcomp,yu2015quality}.
Quartz~\cite{yu2015quality} takes the unusual approach of applying lossy 
compression, by bounding the likelihood that a quality score is informative.
Importantly, Quartz modifies the FASTQ file of the reads and quality scores, and thus can be
readily incorporated into existing sequence analysis pipelines.
Surprisingly, Quartz not only demonstrates improved compression over competing
approaches, but slightly improves the accuracy of downstream variant-calling.

Also taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
similar reads (those that share a common 
substring) together into `buckets', allowing that common substring to now be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This approach allows a general-purpose compressor to achieve tight compression.
SCALCE~\cite{berger2013computational} also relies on a `boosting' scheme, 
reordering
reads in such a way that a general-purpose compressor achieves improved
compression ratios.

@@

\section{Acknowledgements}
This work is supported by the National Institutes of Health, under
grant GM108348.
Y.W.Y. is supported by a Hertz Fellowship.

### 2015-07-31 supplement.tex
final tweaks for submission

\documentclass[article]

@@

As mentioned earlier, function follows form, so in addition to
sequence and expression, structure plays an important role in biological data science.
Protein structure data, primarily determined through X-ray crystallography or
nuclear magnetic resonance (NMR) spectroscopy, describes the coordinates of 
each atom, in Angstroms, of a protein in three-dimensional space.
This information is important to understanding how a protein functions, because
proteins interact physically with other molecules and are geometrically constrained.
Protein structure data are commonly stored in the Protein Data Bank (PDB); 
each PDB entry has an associated resolution indicating the accuracy of the
structural description.
Thus a PDB entry at its most basic level contains a list of atoms, along with
which amino acid they belong to and their spatial coordinates as floating-point
numbers.

>>>>

\documentclass[12pt]{article}

\title{Computational biology in the 21st century:\\Algorithms that scale\\Supplement}
\date{}

@@

\maketitle

\section{Gene expression}

While sequence is important, scientists are also interested in measuring gene expression, or the quantity of
a particular gene product (usually a protein or more recently RNA) present in the cell.
Gene expression data is useful for relating genotype to phenotype, and is 
important in studying many diseases.
Usually obtained through microarrays or RNA-seq technologies, expression data is
quantitative, as each gene from a sample is associated with a numeric expression
level, and high-dimensional, as many thousands of genes may be analyzed at a
time.
Resources such as NCBI’s Gene Expression Omnibus (GEO), which pool 
together many diverse expression studies, now allow researchers to analyze 
thousands of samples generated by others to obtain meaningful biological 
insights. 
More importantly, many of these biological insights can only be gleaned when 
viewed in the context of hundreds or thousands of gene expression samples~\cite{berger2013computational}.

\section{Protein structure}

Protein structure data, primarily determined through X-ray crystallography or
nuclear magnetic resonance (NMR) spectroscopy, describes the coordinates of 
each atom, in Angstroms, of a protein in three-dimensional space.
This information is important to understanding how a protein functions, because
proteins interact physically with other molecules and are geometrically constrained.
Protein structure data are commonly stored in the Protein Data Bank (PDB); 
each PDB entry has an associated resolution indicating the accuracy of the
structural description.
Thus a PDB entry at its most basic level contains a list of atoms, along with
which amino acid they belong to and their spatial coordinates as floating-point
numbers.
